---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Load required library
```{r}
library(dplyr)
```


## Loading and preprocessing the data
  
- Load the concerned dataset into 'activity'  
- Convert 'date' into R 'date' format (observe the difference between activity$date datatype before & after conversion) 

```{r}
activity<-read.csv("./activity.csv", header = TRUE)
str(activity)
activity[,2] <- as.Date(activity$date)
str(activity)
```

- dataset size is 17568 x 3  
- date: 
    - 61 days, 
    - start date on 2012-10-01 (Mon), 
    - end date on 2012-11-30 (Fri), 
    - 8 weekends, 
    - 288 observations on each day
- interval: 
    - minutes value at the interval of 5 mins
    - but minutes value at n\*60 is surprisingly n\*100
    - minutes value resets to 0 on each new day
```{r}
# dataset size is 17568 x 3
dim(activity)

# 61 consecutive days
length(unique(activity$date))

# 288 observations on each day
head(table(activity$date))

# consecutive difference between interval is mostly 5 except at n*60 & on new day
table(diff(activity$interval, 1))
```

## Imputing missing values (moving this question above as this is crucial in identifying no-activity-days)

- NA is found in 'steps' variable only
    - total count of NA is 2304 (8 days of no activity * 288 observations per day)
    - there was no activity captured on 8 of the 61 days
    - 3 options (1. remove those observations, 2. replace NA with 0, 3. replace NA with mean of valid 'steps' in all 'interval' across valid days)
    - Going ahead with 2nd option (reapace NA with 0)
  
```{r}
# NA count in 'steps' column is 2304
sum(is.na(activity$steps))
sum(is.na(activity))

# NA count on each of the days
statsNA <- activity %>% group_by(date) %>% summarise (daysStepsNA = sum(is.na(steps)))
unique(statsNA$daysStepsNA)

# dates on which there was no activity
statsNA$date[which(statsNA$daysStepsNA == 288)]

# Make a new dataset by replacing all NA with 0
activity_NAs_Zero <- activity
activity_NAs_Zero[is.na(activity_NAs_Zero)] <- 0
```

## What is mean total number of steps taken per day?

- Calculate the total number of steps taken per day
    - Both calculations below should give same result  
```{r}
stepsCount <- with(activity, tapply(steps, date, sum, na.rm=TRUE))
stepsCount_NAs_Zero <- with(activity_NAs_Zero, tapply(steps, date, sum))
```

- Make a histogram of the total number of steps taken each day
    - Min & max steps count are 41 & 21194
    - Histogram of steps count on each days with 100 bins which shows 10 instances of near-zero values which is due to 8 days of no-activity and 2 small values (41 & 126 are small w.r.t. max 21194)
```{r}
# min & max steps count
min(stepsCount)
max(stepsCount)

# Hist plot with 100 bins
hist(stepsCount, breaks=100)
```

- Calculate and report the mean and median of the total number of steps taken per day
```{r}
# mean & median of daily steps count incuding no activity days
mean(stepsCount)
median(stepsCount)
```


## What is the average daily activity pattern?

- Make a time series plot (type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis), i.e., plot average of steps count in each intervals
- average daily activity in each interval
    - Both steps mean have been plotted together

```{r}
# Ignore those 8 no acitivity days
stepsSplit <- split(activity$steps, activity$date)
stepsSplit = data.frame(stepsSplit)
dim(stepsSplit)
stepsMean = apply(stepsSplit, 1, mean, na.rm=TRUE)

# Plot steps mean in each interval across days
plot(activity$interval[1:288], stepsMean, type = "l", col="red", xlab = "Intervals", ylab = "Average Steps", main = "Steps mean in intervals over days (original -vs- imputed)")

# Include those 8 no activity days using imputed data
# "na.rm=TRUE" is redundant in this case
stepsSplit_NAs_Zero <- split(activity_NAs_Zero$steps, activity_NAs_Zero$date)
stepsSplit_NAs_Zero = data.frame(stepsSplit_NAs_Zero)
dim(stepsSplit_NAs_Zero)
stepsMean_NAs_Zero = apply(stepsSplit_NAs_Zero, 1, mean, na.rm=TRUE)

# Compare these means
head(data.frame(stepsMean, stepsMean_NAs_Zero))

# Plot steps mean in each interval across days for imputed dataset
lines(activity$interval[1:288], stepsMean_NAs_Zero, col="blue")

```

- Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
activity$interval[which(stepsMean == max(stepsMean))]
```

## Are there differences in activity patterns between weekdays and weekends?

- Split dataset into weekday (53 days) & weekend (8 days)
```{r}
# Mutate dataset with a new column for 'day'
activity_NAs_Zero <- mutate(activity_NAs_Zero, day = ifelse(weekdays(activity_NAs_Zero$date) == "Saturday" | weekdays(activity_NAs_Zero$date) == "Sunday", "weekend", "weekday"))
activity_NAs_Zero$day <- as.factor(activity_NAs_Zero$day)
str(activity_NAs_Zero)

#  Split dataset into weekday (53 days) & weekend (8 days)
activityWeekend <- subset(activity_NAs_Zero, as.character(activity_NAs_Zero$day) == "weekend")
activityWeekday <- subset(activity_NAs_Zero, as.character(activity_NAs_Zero$day) == "weekday")

stepsSplitWeekday <- split(activityWeekday$steps, activityWeekday$date)
stepsSplitWeekday = data.frame(stepsSplitWeekday)
dim(stepsSplitWeekday)
stepsMeanWeekday = apply(stepsSplitWeekday, 1, mean, na.rm=TRUE)

stepsSplitWeekend <- split(activityWeekend$steps, activityWeekend$date)
stepsSplitWeekend = data.frame(stepsSplitWeekend)
dim(stepsSplitWeekend)
stepsMeanWeekend = apply(stepsSplitWeekend, 1, mean, na.rm=TRUE)

par(mfrow=c(2,1))

# Plot steps mean in each interval across days (weekday)
plot(activity$interval[1:288], stepsMeanWeekday, type = "l", xlab = "Intervals", ylab = "Average Steps", main = "Steps mean in each interval across days (weekday)")

# Plot steps mean in each interval across days (weekend)
plot(activity$interval[1:288], stepsMeanWeekend, type = "l", xlab = "Intervals", ylab = "Average Steps", main = "Steps mean in each interval across days (weekend)")
```

